Enhancing Adversarial Robustness for Deep Metric Learning

### 2021-11-13 robdml.tex
import vishal's changes, part 2 (end)

Despite the advancements in this field thanks to deep learning, recent studies
find DML models vulnerable to adversarial attack, where an imperceptible
perturbation can incur unexpected retrieval result, or covertly change the
rankings~\cite{advrank,advorder}.
%
Such vulnerability raises security, safety, and fairness concerns in DML
applications.
%
For example, impersonation or recognition evasion are possible on a vulnerable
DML-based face-identification system.
%
To counter the attacks (\ie, mitigating the vulnerability), the
\emph{adversarial robustness} of DML models has to be improved via defense.

\begin{figure}[t]
	\includegraphics[width=1.0\columnwidth]{introplot.pdf}
	%\vspace{-2.0em}
	\caption{
		%
		Comparison in Robustness, Training Cost, and Recall@1 
		between Our Method (\ie, ``HM[$\mathcal{S},g_\mathsf{LGA}$]\&ICS'')
		and the State-of-The-Art Method (\ie, ``ACT[$\mathcal{R}$]'' and
		``ACT[$\mathcal{S}$]'') on CUB Dataset.
		%
	}
	\label{fig:introplot}
\end{figure}

@@

Previous works imply that an appropriate adversary for the inner
\emph{maximization} problem should increase the loss to an ``intermediate''
point between that of benign examples (\ie, unperturbed examples) and the
theoretical upper bound.
%
In this paper, we argue such point can be reached by a direct and efficient
adversary.
%
Besides, we conjecture the triplet sampling strategy has a key impact in
adversarial training, because it is also able to greatly influence the
mathematical expectation of loss even without adversarial attack.

@@

Then, how hard should such ``\emph{destination} hardness'' be?
%
Recall that the model is already prone to collapse with excessively hard benign
triplets~\cite{facenet}, let alone adversarial examples.
%
Thus, intuitively, the \emph{destination} hardness can be the hardness of
another benign triplet which is moderately harder than the given triplet (\eg,
a Semihard~\cite{facenet} triplet).
%
However, in the late phase of training, the expectation of the difference
between such \emph{destination} hardness and that of the given triplet will be
small, leading to weak adversarial examples and inefficient adversarial
learning.
%
Besides, strong adversarial examples in the early phase of training may also
hinder the model from learning good embeddings, and hence influence the
performance on benign examples.
%
Namely, a better \emph{destination} hardness should be able to balance the
training objectives in the early and late phases of training.

@@

To this end, Graduate Adversary, a family of pseudo-hardness functions is
proposed, which can be used as the \emph{destination} hardness.
%
A function that leads to relatively weak and relatively strong adversarial
examples respectively in the early and late phase of training belongs to this
family.
%
As an example, we design a ``Linear Graduate Adversary'' (LGA)
function as the linearly scaled negative triplet margin, incorporating a strong
prior that the \emph{destination} hardness should remain Semihard
based on our empirical observation.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{hmflexible.pdf}
	\vspace{-1.8em}
	\caption{Flexibility of Hardness Manipulation.
	%
	Regular training and min-max adversarial training are its boundary cases.
	}
	\label{fig:hmflexible}
\end{figure}

% contributions
In brief, our contributions include proposing:
%
\begin{enumerate}[nosep, noitemsep, leftmargin=*]
	%
	\item {\textit{Hardness Manipulation}} (HM) as a flexible and efficient
		tool to create adversarial example triplets for subsequent adversarial
		training of DML model.
		%
	\item \textit{Linear Gradual Adversary} (LGA) as a Graduate Adversary, \ie,
		a pseudo-hardness function for HM, which incorporates our empirical
		observations and can balance the training objectives during the
		training process.
		%
	\item \textit{Intra-Class Structure} (ICS) loss term to further improve
		model robustness and adversarial training efficiency, while such structure is
		neglected by existing methods. 
		%
\end{enumerate}

@@

\textbf{Adversarial Defense.}
%
To battle against the attacks, some defenses incur gradient masking
effect, but leads to a false sense of security~\cite{obfuscated}.
%
Defensive distillation~\cite{distill2} is compromised~\cite{cw}.
%
Ensemble of weak defenses is not robust~\cite{ensembleweak}.
%
Other defenses are also proposed, such as input
preprocessing~\cite{deflecting}, or a randomization inside the
network~\cite{self-ensemble}.
%
But many defenses are still susceptible to adaptive attack~\cite{adaptive}.
%
Of all defenses, adversarial training remains to be one of the most effective
method~\cite{bilateral,advtrain-triplet,benchmarking}, consistently showing
promising empirical robustness, although suffers from high training
cost and performance drop on benign examples~\cite{odds}.

@@

\textbf{Attacks in DML.}
%
DML has been found vulnerable to adversarial attacks as
well~\cite{advrank,advorder,robrank}, which raises concerns on safety,
security, or fairness for a DML application.
%
The existing attacks aims to completely subvert the image retrieval
results~\cite{qair,learn-to-misrank,advdpqn,advpattern,flowertower,universalret},
or covertly alter the top-ranking results without being
abnormal~\cite{advrank,advorder}.

%
\textbf{Defenses in DML.} Different from attacks, defense methods are less
explored.
%
Embedding Shifted Triplet (EST)~\cite{advrank} is an
adversarial training method using adversarial examples with maximized embedding
move distance off their original locations.
%
Anti-Collapse Triplet (ACT)~\cite{robrank} forces the model to separate
collapsed positive and negative samples apart in order to learn robust
features, which achieves the state-of-the-art robustness.
%
However, the existing defense methods suffer from low efficiency as the
inner problem is replaced into a indirect form in order to avoid model collapse.

@@

In Deep Metric Learning (DML)~\cite{revisiting,dmlreality}, an embedding
function $\phi:\mathcal{X}\mapsto \Phi \subseteq \mathbb{R}^D$ is learned to
map data points $\mX\in\mathcal{X}$ into an embedding space $\Phi$, which is usually
normalized to the real unit hypersphere for regularization.
%
With a predefined distance function $d(\cdot,\cdot)$, which is usually the
Euclidean distance, we can measure the distance between $\mX_i$ and $\mX_j$ as
$d_\phi(\mX_i,\mX_j)=d(\phi(\mX_i),\phi(\mX_j))$.
%
In a typical setting, the triplet loss~\cite{facenet} can be used to learn the
embedding function, and it could reach the state-of-the-art performance with an
appropriate triplet sampling strategy~\cite{revisiting}.

@@

Given a sample triplet of anchor, positive, negative images, \ie, $\mA, \mP,
\mN \in \mathcal{X}$, we first calculate their embeddings with $\phi(\cdot)$ as
$\va, \vp, \vn$, respectively.
%
Then the triplet loss~\cite{facenet} is defined as:
%
$
%
	L_\text{T}(\va, \vp, \vn; \gamma) = \max(0, d(\va, \vp) - d(\va, \vn) +
	\gamma),
%
$
%
where $\gamma$ is a predefined margin parameter.
%
To attack the DML model, an imperceptible adversarial perturbation $\vr \in
\Gamma$ is added to the input image $\mX$, where $\Gamma = \{\vr | \mX+\vr\in
\mathcal{X},  \|\vr\|_p \leq \varepsilon\}$, so that its embedding vector
$\tilde{\vx}=\phi(\mX+\vr)$ will be moved off its original location towards
other positions to achieve the attacker's goal.
%
To defend against the attacks, the DML model can be adversarially trained to
reduce the effect of attacks~\cite{advrank,robrank}.
%
The most important metrics for a good defense is training cost, adversarial
robustness, and performance drop on benign examples.

@@

Then, \emph{Hardness Manipulation} (HM) aims to increase the \emph{source}
hardness $H_\mathsf{S}$ into a specified ``\emph{destination} hardness''
$H_\mathsf{D}$, by finding adversarial examples of the source triplet, \ie,
$(\mA{+} {\vr}_a, \mP {+} {\vr}_p, \mN {+} {\vr}_n)$, where $({\vr}_a, {\vr}_p,
{\vr}_n)$ are the adversarial perturbations.
%
Denoting the hardness of the adversarially perturbed \emph{source} triplet
as $\tilde{H}_\mathsf{S}$, \ie,
%
$\tilde{H}_\mathsf{S} = H(\mA{+} {\vr}_a, \mP {+} {\vr}_p, \mN {+} {\vr}_n)$,
%
the HM is implemented as:
%
\begin{equation}
	%
	\hat{\vr}_a, \hat{\vr}_p, \hat{\vr}_n = \argmin_{\vr_a,\vr_p,\vr_n}
	\big\|\max(0, H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \big\|_2^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\max(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when
$H_\mathsf{S}>H_\mathsf{D}$, automatically stopping the optimization, because
$H_\mathsf{S}$ is not desired to be reduced once it exceeds $H_\mathsf{D}$.
%
The \cref{eq:hm} is written in the L-$2$ norm form instead of the standard Mean
Squared Error because HM can be directly extended into vector form for a
mini-batch.
%
The optimization problem can be solved by Projected Gradient Descent
(PGD)~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model with
%
$L_\text{T}(\phi(\mA+\hat{\vr}_a), \phi(\mP+\hat{\vr}_p),
\phi(\mN+\hat{\vr}_n))$.
%
The overall procedure of HM is illustrated in \cref{fig:hm}.
%
For convenience,
we abbreviate the adversarial training with adversarial examples created
through this way as $\text{HM}[H_\mathsf{S},H_\mathsf{D}]$ in this paper.

\begin{figure}
	\includegraphics[width=\columnwidth]{hmillust.pdf}
	\vspace{-1.8em}
	\caption{Illustration of Hardness Manipulation.}
	\label{fig:hm}
\end{figure}

@@

If we have a strong prior knowledge on what the \emph{destination} hardness
should be, we can even use a pseudo-hardness function $g(\cdot)$, \ie, a
customized scalar function.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{gaillust.pdf}
	\caption{Illustration of (Linear) Gradual Adversary.}
	\label{fig:ga}
\end{figure}

@@

Intuitively, such deficiency can be alleviated with a \emph{pseudo-hardness}
function that slightly increase the value of $H_\mathsf{D}$ in the late phase
of training.
%
Denoting the loss value from the previous training iteration as $\ell_{t-1}$,
we first normalize it into $[0,1]$ as $\bar{\ell}_{t-1}=\min(u,\ell_{t-1})/u$,
where $u$ is a manually specified constant.
%
Then we can linearly shift the $E[H_\mathsf{D}]$ by a scaled constant $\xi$,
\ie,
%
\begin{equation}
	%
	g_\mathsf{B}(H_\mathsf{D};\xi,\bar{\ell}_{t-1}) =
	H_\mathsf{D} + \xi \cdot (1-\bar{\ell}_{t-1}).
	%
\end{equation}
%
The deficiency can be alleviated in
$\text{HM}[H_\mathsf{S},g_\mathsf{B}(H_\mathsf{D})]$.

@@

As an example, we propose a ``Linear Gradual Adversary'' (LGA) pseudo-hardness
function that is independent to any benign triplets, incorporating an empirical
observation that $H_\mathsf{D}$ should remain Semihard~\cite{facenet}, as
follows:
%
\begin{equation}
	%
	g_\mathsf{LGA}(\bar{\ell}_{t-1}) = -\gamma \cdot \bar{\ell}_{t-1} ~ \in
	[-\gamma,0].
	%
\end{equation}
%
And the training objectives, namely performance on benign examples and
robustness will be automatically balanced in
$\text{HM}[H_\mathsf{S},g_\mathsf{LGA}]$, leading to a better eventual overall
performance.
%
More complicated or non-linear pseudo-hardness functions are left for future study
for sake of simplicity.
%

@@

\begin{figure}[t]
	\includegraphics[width=1.0\columnwidth]{icsapn.pdf}
	\caption{Illustration of Intra-Class Structure Loss Term.}
	\label{fig:ics}
\end{figure}

@@

To validate the proposed defense, we conduct experiments
on three commonly used DML datasets: CUB-200-2011 (CUB)~\cite{cub200}, Cars-196
(CARS)~\cite{cars196}, and Stanford-Online-Products (SOP)~\cite{sop}.
%
We follow the same experiment setup as the state-of-the-art work~\cite{robrank,revisiting}
for ease of comparison.

@@

Specifically, we (adversarially) train ImageNet-initialized ResNet-18
(RN18)~\cite{resnet} with the output dimension of the last layer changed to
$N{=}512$.
%
The margin $\gamma$ in triplet loss is $0.2$, which is consistent to standard
DML~\cite{revisiting}.
%
Adam~\cite{adam} optimizer is used for parameter updates, with a learning rate
of $1.0{\times}10^{-3}$ for $150$ epochs, and a mini-batch size of $112$.
%
Adversarial examples are created under the perturbation budget
$\varepsilon{=}8/255$ in infinity norm, using PGD~\cite{madry} with step size
$\alpha{=}1/255$ and a default maximum step number $\eta{=}8$.
%
The parameter $u$ is equal to $\gamma$, much less than the loss upper bound
in order to avoid excessive hardness boost in $g_\mathsf{B}$ and $g_\mathsf{LGA}$.
%
Parameter $\lambda$ for $L_\text{ICS}$ is $0.5$ by default ($0.05$ on SOP).

The model performance on benign (\ie, unperturbed) examples is measured in
Recall@1 (R@1), Recall@2 (R@2), mAP and NMI
following~\cite{revisiting,robrank}.
%
The adversarial robustness of a model is measured in Empirical Robustness Score
(ERS)~\cite{robrank}, a normalized score (the higher the better) from
an collection of existing attacks against DML or their simplified white-box
alternatives optimized with PGD ($\eta=32$ for strong attack).
%
Since adversarial training is not ``gradient masking''~\cite{obfuscated}, the
performance of white-box attack can be regarded as the upper bound of the
black-box attacks, and thus a model that is empirically robust to the collection
of white-box attacks is expected to be robust in general.

Concretely, the collection of attacks for ERS includes:
%
(1) CA+, CA-, QA+ and QA-~\cite{advrank}, which move some selected candidates
towards the topmost or bottommost part of ranking list;
%
(2) TMA~\cite{flowertower} which increases the cosine similarity between two selected samples;
%
(3) ES~\cite{advrank,advdpqn}, which moves the embedding of a sample off its original position as
distant as possible;
%
(4) LTM~\cite{learn-to-misrank}, which perturbs the ranking result by minimizing the distance of
unmatched pairs while maximizing the distance of matched pairs;
%
(5) GTM~\cite{robrank}, which minimizes the distance between query and the
closest unmatching sample.
%
(6) GTT~\cite{robrank}, aims to move the top-$1$ candidate out of the top-$4$
retrieval results, which is simplified from \cite{qair}.
%
The setup of all the attacks for robustness evaluation are unchanged from
\cite{robrank} for fair comparison.
%
Further details can be found in \cite{robrank} or the supplementary.

@@

HM is flexible so that any existing or future triplet sampling strategy
can be used for the source triplet or calculating $H_\mathsf{D}$.
%
But not all potential combinations are expected to be effective for HM, as it
will not create an adversarial triplet when $H_\mathsf{S}\geqslant
H_\mathsf{D}$.
%
Thus, we sort the strategies based on the mean hardness of their
outputs in~\cref{tab:hsort}.
%
Then we adversarially train models on CUB dataset with all combinations
respectively, and summarize their R@1 and ERS in \cref{tab:desth}.

@@

In contrast, HM[$\mathcal{R},\mathcal{M}$] achieves an even higher ERS under
the same training cost, but with a larger penalty in R@1 compared to
ACT[$\mathcal{R}$].
%
Compared to ACT[$\mathcal{S}$], HM[$\mathcal{S},\mathcal{M}$] is able to retain
a relatively high R@1, but in a much higher efficiency.
%
According to \cref{fig:hmeff}, HM[$\mathcal{R},\mathcal{M}$] achieves the
highest ERS and efficiency but with the most significant drop in R@1, which is
not acceptable in applications.
%
Apart from that, HM[$\mathcal{S},\mathcal{M}$] achieves a promising result
in every aspect.
%
Its efficiency in gaining robustness is basically on par with
ACT[$\mathcal{R}$], but can achieve a significantly higher R@1.
%
It achieves a balance between ERS and R@1 on par with ACT[$\mathcal{S}$], but
in a significantly higher efficiency.

@@

\noindent\textbf{II. Gradual Adversary.}
%
As discussed in \cref{sec:32}, the HM[$\mathcal{S},\mathcal{M}$] may still
suffer from the imbalance between learning the embeddings and gaining
adversarial robustness.
%
Hence, we conduct further experiments %with an improved $H_\mathsf{D}$
following the discussion, as shown in \cref{tab:gaeff} and \cref{fig:gaeff}.
%
Compared to HM[$\mathcal{S},\mathcal{M}$], slightly boosting the hardness
training with $g_\mathsf{B}(\cdot)$ where $\xi{=}0.1$ benefits the ERS, but
results in a notably lower R@1;
%
A constant $H_\mathsf{D}$ at the upper bound of Semihard (\ie, $0$; too high
for both the early and the late phase of training) renders model collapse;
%
$H_\mathsf{D}$ at the lower bound (\ie, $-\gamma$; too low for the late phase)
leads to insignificant ERS improvement;
%
$H_\mathsf{D}{=}-\gamma/2$ provides a balance in training objectives, but
still suffers from inflexibility.
%
In contrast, being not susceptible to the mentioned problems of other choices,
HM[$\mathcal{S},g_\mathsf{LGA}$] achieves an ERS on par with
HM[$\mathcal{S},\mathcal{M}$], but is at the lowest R@1 performance penalty among all choices.

@@

Further experiment discussions, technical details, limitations and societal
impact are presented in supplementary.

>>>>

Despite the advancements in this field thanks to deep learning, recent studies
find DML models vulnerable to adversarial attacks, where  imperceptible
perturbations can incur unexpected retrieval result, or covertly change the
rankings~\cite{advrank,advorder}.
%
Such vulnerability raises security, safety, and fairness concerns in the DML
applications.
%
For example, impersonation or recognition evasion are possible on a vulnerable
DML-based face-identification system.
%
To counter the attacks (\ie, mitigating the vulnerability), the
\emph{adversarial robustness} of DML models has to be improved via defense.

\begin{figure}[t]
	\includegraphics[width=1.0\columnwidth]{introplot.pdf}
	%\vspace{-2.0em}
	\caption{
		%
		Comparison in robustness, training cost, and recall@1 
		between our method (\ie, ``HM[$\mathcal{S},g_\mathsf{LGA}$]\&ICS'')
		and the state-of-the-art method (\ie, ``ACT[$\mathcal{R}$]'' and
		``ACT[$\mathcal{S}$]'') on the CUB Dataset.
		%
	}
	\label{fig:introplot}
\end{figure}

@@

{\bf{Previous works \cite{} imply that an appropriate adversary for the inner
\emph{maximization} problem should increase the loss to an ``intermediate''
point between that of benign examples (\ie, unperturbed examples) and the
theoretical upper bound.}}
%
In this paper, we argue such point can be reached by a direct and efficient
adversary.
%
Besides, we conjecture the triplet sampling strategy has a key impact in
adversarial training, because it is also able to greatly influence the
mathematical expectation of loss even without adversarial attack.

@@

Then, how hard should such ``\emph{destination} hardness'' be?
%
Recall that the model is already prone to collapse with excessively hard benign
triplets~\cite{facenet}, let alone adversarial examples.
%
Thus, intuitively, the \emph{destination} hardness can be the hardness of
another benign triplet which is moderately harder than the given triplet (\eg,
a Semihard~\cite{facenet} triplet).
%
However, in the late phase of training, the expectation of the difference
between such \emph{destination} hardness and that of the given triplet will be
small, leading to weak adversarial examples and inefficient adversarial
learning.
%
Besides, strong adversarial examples in the early phase of training may also
hinder the model from learning good embeddings, and hence influence the
performance on benign examples.
%
In particular, a better \emph{destination} hardness should be able to balance the
training objectives in the early and late phases of training.

@@

To this end, Graduate Adversary, a family of pseudo-hardness functions is
proposed, which can be used as the \emph{destination} hardness.
%
A function that leads to relatively weak and relatively strong adversarial
examples, respectively in the early and late phase of training belongs to this
family.
%
As an example, we design a ``Linear Graduate Adversary'' (LGA)
function as the linearly scaled negative triplet margin, incorporating a strong
prior that the \emph{destination} hardness should remain Semihard
based on our empirical observation.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{hmflexible.pdf}
	\vspace{-1.8em}
	\caption{Flexibility of hardness manipulation.
	%
	Regular training and min-max adversarial training are its boundary cases.
	}
	\label{fig:hmflexible}
\end{figure}

% contributions
In summary, our contributions include proposing:
%
\begin{enumerate}[nosep, noitemsep, leftmargin=*]
	%
	\item {\textit{Hardness Manipulation}} (HM) as a flexible and efficient
		tool to create adversarial example triplets for subsequent adversarial
		training of a DML model.
		%
	\item \textit{Linear Gradual Adversary} (LGA) as a Graduate Adversary, \ie,
		a pseudo-hardness function for HM, which incorporates our empirical
		observations and can balance the training objectives during the
		training process.
		%
	\item \textit{Intra-Class Structure} (ICS) loss term to further improve
		model robustness and adversarial training efficiency, while such structure is
		neglected by existing methods. 
		%
\end{enumerate}

@@

\textbf{Adversarial Defense.}
%
To battle against the attacks, some defenses incur gradient masking
effect, but often lead to a false sense of security~\cite{obfuscated}.
%
Defensive distillation~\cite{distill2} is compromised in~\cite{cw}.
%
It has been shown that ensemble of weak defenses is not robust~\cite{ensembleweak}.
%
Other defenses are also proposed, such as input
preprocessing~\cite{deflecting}, or a randomization inside the
network~\cite{self-ensemble}.
%
But many defenses are still susceptible to adaptive attacks~\cite{adaptive}.
%
Of all defenses, adversarial training remains to be one of the most effective
methods~\cite{bilateral,advtrain-triplet,benchmarking}, consistently showing
promising empirical robustness, although suffers from high training
cost and performance drop on benign examples~\cite{odds}.

@@

\textbf{Attacks in DML.}
%
DML has been found vulnerable to adversarial attacks as
well~\cite{advrank,advorder,robrank}, which raises concerns on safety,
security, or fairness for a DML application.
%
The existing attacks aim to completely subvert the image retrieval
results~\cite{qair,learn-to-misrank,advdpqn,advpattern,flowertower,universalret},
or covertly alter the top-ranking results without being
abnormal~\cite{advrank,advorder}.

%
\textbf{Defenses in DML.} Different from attacks, defense methods are less
explored.
%
Embedding Shifted Triplet (EST)~\cite{advrank} is an
adversarial training method using adversarial examples with maximized embedding
move distance off their original locations.
%
Anti-Collapse Triplet (ACT)~\cite{robrank} forces the model to separate
collapsed positive and negative samples apart in order to learn robust
features, which achieves the state-of-the-art robustness.
%
However, the existing defense methods suffer from low efficiency as the
inner problem is replaced into an indirect form in order to avoid model collapse.

@@

In DML~\cite{revisiting,dmlreality}, an embedding
function $\phi:\mathcal{X}\mapsto \Phi \subseteq \mathbb{R}^D$ is learned to
map data points $\mX\in\mathcal{X}$ into an embedding space $\Phi$, which is usually
normalized to the real unit hypersphere for regularization.
%
With a predefined distance function $d(\cdot,\cdot)$, which is usually the
Euclidean distance, we can measure the distance between $\mX_i$ and $\mX_j$ as
$d_\phi(\mX_i,\mX_j)=d(\phi(\mX_i),\phi(\mX_j))$.
%
In a typical setting, the triplet loss~\cite{facenet} can be used to learn the
embedding function, and it could reach the state-of-the-art performance with an
appropriate triplet sampling strategy~\cite{revisiting}.

@@

Given a sample triplet of anchor, positive, negative images, \ie, $\mA, \mP,
\mN \in \mathcal{X}$, we first calculate their embeddings with $\phi(\cdot)$ as
$\va, \vp, \vn$, respectively.
%
Then the triplet loss~\cite{facenet} is defined as:
\setlength{\belowdisplayskip}{0pt} \setlength{\belowdisplayshortskip}{0pt}
\setlength{\abovedisplayskip}{0pt} \setlength{\abovedisplayshortskip}{0pt}
%
\begin{equation}
%
	L_\text{T}(\va, \vp, \vn; \gamma) = \max(0, d(\va, \vp) - d(\va, \vn) +
	\gamma),
	\end{equation}
%
%$
%
where $\gamma$ is a predefined margin parameter.
%
To attack the DML model, an imperceptible adversarial perturbation $\vr \in
\Gamma$ is added to the input image $\mX$, where $\Gamma = \{\vr | \mX+\vr\in
\mathcal{X},  \|\vr\|_p \leq \varepsilon\}$, so that its embedding vector
$\tilde{\vx}=\phi(\mX+\vr)$ will be moved off its original location towards
other positions to achieve the attacker's goal.
%
To defend against the attacks, the DML model can be adversarially trained to
reduce the effect of attacks~\cite{advrank,robrank}.
%
The most important metrics for a good defense is training cost, adversarial
robustness, and performance drop on benign examples.

@@

Then, \emph{Hardness Manipulation} (HM) aims to increase the \emph{source}
hardness $H_\mathsf{S}$ into a specified ``\emph{destination} hardness''
$H_\mathsf{D}$, by finding adversarial examples of the source triplet, \ie,
$(\mA{+} {\vr}_a, \mP {+} {\vr}_p, \mN {+} {\vr}_n)$, where $({\vr}_a, {\vr}_p,
{\vr}_n)$ are the adversarial perturbations.
%
Denoting the hardness of the adversarially perturbed \emph{source} triplet
as $\tilde{H}_\mathsf{S}$, \ie,
%
$\tilde{H}_\mathsf{S} = H(\mA{+} {\vr}_a, \mP {+} {\vr}_p, \mN {+} {\vr}_n)$,
%
the HM is implemented as:
%
\begin{equation}
	%
	\hat{\vr}_a, \hat{\vr}_p, \hat{\vr}_n = \argmin_{\vr_a,\vr_p,\vr_n}
	\big\|\max(0, H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \big\|_2^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\max(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when
$H_\mathsf{S}>H_\mathsf{D}$, automatically stopping the optimization, because
$H_\mathsf{S}$ is not desired to be reduced once it exceeds $H_\mathsf{D}$.
%
 \cref{eq:hm} is written in the L-$2$ norm form instead of the standard Mean
Squared Error because HM can be directly extended into vector form for a
mini-batch.
%
The optimization problem can be solved by Projected Gradient Descent
(PGD)~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model with
%
$L_\text{T}(\phi(\mA+\hat{\vr}_a), \phi(\mP+\hat{\vr}_p),
\phi(\mN+\hat{\vr}_n))$.
%
The overall procedure of HM is illustrated in \cref{fig:hm}.
%
For convenience,
we abbreviate the adversarial training with adversarial examples created
through this way as $\text{HM}[H_\mathsf{S},H_\mathsf{D}]$ in this paper.

\begin{figure}
	\includegraphics[width=\columnwidth]{hmillust.pdf}
	\vspace{-1.8em}
	\caption{Illustration of hardness manipulation.}
	\label{fig:hm}
\end{figure}

@@

If we have a strong prior knowledge on what the \emph{destination} hardness
should be, then we can even use a pseudo-hardness function $g(\cdot)$, \ie, a
customized scalar function.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{gaillust.pdf}
	\caption{Illustration of (linear) gradual adversary.}
	\label{fig:ga}
\end{figure}

@@

Intuitively, such deficiency can be alleviated with a \emph{pseudo-hardness}
function that slightly increases the value of $H_\mathsf{D}$ in the late phase
of training.
%
Denoting the loss value from the previous training iteration as $\ell_{t-1}$,
we first normalize it into $[0,1]$ as $\bar{\ell}_{t-1}=\min(u,\ell_{t-1})/u$,
where $u$ is a manually specified constant.
%
Then we can linearly shift the $E[H_\mathsf{D}]$ by a scaled constant $\xi$,
\ie,
%
\begin{equation}
	%
	g_\mathsf{B}(H_\mathsf{D};\xi,\bar{\ell}_{t-1}) =
	H_\mathsf{D} + \xi \cdot (1-\bar{\ell}_{t-1}).
	%
\end{equation}
%
The deficiency can be alleviated in
$\text{HM}[H_\mathsf{S},g_\mathsf{B}(H_\mathsf{D})]$.

@@

As an example, we propose a ``Linear Gradual Adversary'' (LGA) pseudo-hardness
function that is independent to any benign triplets, incorporating an empirical
observation that $H_\mathsf{D}$ should remain Semihard~\cite{facenet}, as
follows:
%
\begin{equation}
	%
	g_\mathsf{LGA}(\bar{\ell}_{t-1}) = -\gamma \cdot \bar{\ell}_{t-1} ~ \in
	[-\gamma,0].
	%
\end{equation}
%
And the training objectives, namely performance on benign examples and
robustness will be automatically balanced in
$\text{HM}[H_\mathsf{S},g_\mathsf{LGA}]$, leading to a better eventual overall
performance.
%
More complicated or non-linear pseudo-hardness functions are left for future study.
%for sake of simplicity.
%

@@

\begin{figure}[t]
	\includegraphics[width=1.0\columnwidth]{icsapn.pdf}
	\caption{Illustration of intra-class structure loss term.}
	\label{fig:ics}
\end{figure}

@@

To validate the proposed defense, we conduct experiments
on three commonly used DML datasets: CUB-200-2011 (CUB)~\cite{cub200}, Cars-196
(CARS)~\cite{cars196}, and Stanford-Online-Products (SOP)~\cite{sop}.
%
We follow the same experimental setup as that used in the state-of-the-art works~\cite{robrank,revisiting}
for ease of comparison.

@@

Specifically, we (adversarially) train ImageNet-initialized ResNet-18
(RN18)~\cite{resnet} with the output dimension of the last layer changed to
$N{=}512$.
%
The margin $\gamma$ in the triplet loss is $0.2$, which is consistent with the standard
DML~\cite{revisiting}.
%
Adam~\cite{adam} optimizer is used for parameter updates, with a learning rate
of $1.0{\times}10^{-3}$ for $150$ epochs, and a mini-batch size of $112$.
%
Adversarial examples are created under the perturbation budget
$\varepsilon{=}8/255$ in infinity norm, using PGD~\cite{madry} with step size
$\alpha{=}1/255$ and a default maximum step number $\eta{=}8$.
%
The parameter $u$ is equal to $\gamma$, much less than the loss upper bound
in order to avoid excessive hardness boost in $g_\mathsf{B}$ and $g_\mathsf{LGA}$.
%
Parameter $\lambda$ for $L_\text{ICS}$ is $0.5$ by default ($0.05$ on SOP).

The model performance on the benign (\ie, unperturbed) examples is measured in terms of 
Recall@1 (R@1), Recall@2 (R@2), mAP and NMI
following~\cite{revisiting,robrank}.
%
The adversarial robustness of a model is measured in Empirical Robustness Score
(ERS)~\cite{robrank}, a normalized score (the higher the better) from
an collection of existing attacks against DML or their simplified white-box
alternatives optimized with PGD ($\eta=32$ for strong attack).
%
Since adversarial training is not ``gradient masking''~\cite{obfuscated}, the
performance of white-box attack can be regarded as the upper bound of the
black-box attacks, and thus a model that is empirically robust to the collection
of white-box attacks is expected to be robust in general.

Concretely, the collection of attacks for ERS include:
%
(1) CA+, CA-, QA+ and QA-~\cite{advrank}, which move some selected candidates
towards the topmost or bottommost part of ranking list;
%
(2) TMA~\cite{flowertower} which increases the cosine similarity between two selected samples;
%
(3) ES~\cite{advrank,advdpqn}, which moves the embedding of a sample off its original position as
distant as possible;
%
(4) LTM~\cite{learn-to-misrank}, which perturbs the ranking result by minimizing the distance of
unmatched pairs while maximizing the distance of matched pairs;
%
(5) GTM~\cite{robrank}, which minimizes the distance between query and the
closest unmatching sample.
%
(6) GTT~\cite{robrank}, aims to move the top-$1$ candidate out of the top-$4$
retrieval results, which is simplified from \cite{qair}.
%
The setup of all the attacks for robustness evaluation are unchanged from
\cite{robrank} for fair comparison.
%
Further details can be found in \cite{robrank} and in the supplementary document.

@@

HM is flexible so that any existing or future triplet sampling strategy
can be used for the source triplet or calculating $H_\mathsf{D}$.
%
But not all potential combinations are expected to be effective for HM, as it
will not create an adversarial triplet when $H_\mathsf{S}\geqslant
H_\mathsf{D}$.
%
Thus, we sort the strategies based on the mean hardness of their
outputs in~\cref{tab:hsort}.
%
Then we adversarially train models on the CUB dataset with all combinations
respectively, and summarize their R@1 and ERS in \cref{tab:desth}.

@@

In contrast, HM[$\mathcal{R},\mathcal{M}$] achieves an even higher ERS under
the same training cost, but with a larger penalty in R@1 compared to
ACT[$\mathcal{R}$].
%
Compared to ACT[$\mathcal{S}$], HM[$\mathcal{S},\mathcal{M}$] is able to retain
a relatively high R@1, but in a much higher efficiency.
%
As can be seen from \cref{fig:hmeff}, HM[$\mathcal{R},\mathcal{M}$] achieves the
highest ERS and efficiency but with the most significant drop in R@1, which is
not acceptable in applications.
%
Apart from that, HM[$\mathcal{S},\mathcal{M}$] achieves a promising result
in every aspect.
%
Its efficiency in gaining robustness is basically on par with
ACT[$\mathcal{R}$], but can achieve a significantly higher R@1.
%
It achieves a balance between ERS and R@1 on par with ACT[$\mathcal{S}$], but
in a significantly higher efficiency.

@@

\noindent\textbf{II. Gradual Adversary.}
%
As discussed in \cref{sec:32},  HM[$\mathcal{S},\mathcal{M}$] may still
suffer from the imbalance between learning the embeddings and gaining
adversarial robustness.
%
Hence, we conduct further experiments %with an improved $H_\mathsf{D}$
following the discussion, as shown in \cref{tab:gaeff} and \cref{fig:gaeff}.
%
Compared to HM[$\mathcal{S},\mathcal{M}$], slightly boosting the hardness
training with $g_\mathsf{B}(\cdot)$ where $\xi{=}0.1$ benefits the ERS, but
results in a notably lower R@1;
%
A constant $H_\mathsf{D}$ at the upper bound of Semihard (\ie, $0$; too high
for both the early and the late phase of training) renders model collapse;
%
$H_\mathsf{D}$ at the lower bound (\ie, $-\gamma$; too low for the late phase)
leads to insignificant ERS improvement;
%
$H_\mathsf{D}{=}-\gamma/2$ provides a balance in training objectives, but
still suffers from inflexibility.
%
In contrast, being not susceptible to the mentioned problems of other choices,
HM[$\mathcal{S},g_\mathsf{LGA}$] achieves an ERS on par with
HM[$\mathcal{S},\mathcal{M}$], but is at the lowest R@1 performance penalty among all choices.

@@

Further discussions on experiments, technical details, limitations and societal
impact are presented in the supplementary document.
